%!TEX root = thesis.tex

End-to-end autonomous driving models achieve strong performance on aggregate benchmarks but suffer severe degradation when deployed across different cities due to visual domain shift in road geometry, vegetation, signage, and infrastructure. Current benchmarks like NAVSIM mask this failure by aggregating evaluation across cities. This thesis investigates the root cause and demonstrates that cross-city generalization failure is primarily a \emph{representation} problem, not a planning architecture problem.

We make four contributions. First, we introduce \emph{NAVSIM City Splits}, the first per-city train/test partitioning of the NAVSIM benchmark across Boston, Pittsburgh, Las Vegas, and Singapore, enabling systematic cross-city generalization evaluation. Second, through a comprehensive study of 19 backbone configurations spanning three self-supervised learning (SSL) paradigms --- I-JEPA (predictive), DINOv2 (discriminative), and MAE (generative) --- against supervised ImageNet ResNet-34 features, we show that SSL representations reduce cross-city generalization gaps by an order of magnitude across both TransFuser and LAW planning architectures. Third, we integrate SSL backbones into DiffusionDrive, a state-of-the-art truncated diffusion planner, creating both multi-modal and camera-only variants that achieve up to 86.9 PDMS, and evaluate whether the generalization benefits of SSL extend from regression-based to generative planning heads. Fourth, we provide a mechanistic analysis of \emph{why} SSL features generalize better, using Centered Kernel Alignment, linear probing for city classification, and attention visualization to show that SSL produces city-invariant representations that focus on task-relevant road structure rather than city-specific textures. Our findings provide practical guidance for building geographically robust autonomous driving systems: self-supervised pre-training is the single most impactful change for cross-city robustness.
