%!TEX root = thesis.tex

%% NYU Tandon Guidelines for Vita:
%% - Place of birth (do NOT include date of birth)
%% - Brief educational and professional history
%% - Period of time devoted to the research
%% - Laboratories in which research was performed
%% - Source of any special support (grants, fellowships, etc.)
%% - This is NOT a resume

<<<<<<< Updated upstream
Ali Hamza was born in Karachi, Pakistan. He received his Bachelor of Science degree in Computer Science from Habib University in 2022.

% TODO: Fill in educational and professional history

He enrolled at New York University Tandon School of Engineering in 2024 to pursue a Master of Science in Electrical Engineering. His research was conducted under the supervision of Prof.\ Anna Choromanska in the Department of Electrical and Computer Engineering, focusing on self-supervised learning for autonomous driving.

The research presented in this thesis was performed from Fall 2025 to Spring 2026 using the NYU Greene High Performance Computing cluster. 
=======
Ali Hamza was born in Karachi, Pakistan. He received his Bachelor of Science degree in Computer Science from Habib University, Karachi, in May 2022.
He enrolled at New York University Tandon School of Engineering in Fall 2024 to pursue a Master of Science in Computer Engineering. His research was conducted under the supervision of Prof. Anna Choromanska in the Department of Electrical and Computer Engineering, focusing on self-supervised learning for end-to-end trajectory planning in autonomous driving.
The research presented in this thesis was performed from January 2025 to May 2026 using the NYU Greene High Performance Computing cluster. This work was supported by [FUNDING SOURCE, if any].
>>>>>>> Stashed changes
