%!TEX root = ../thesis.tex
\chapter{Methodology}
\label{chap:methodology}

%% Approx. 15 pages.
%% Main thrusts: exploratory lightweight-head SSL, cross-city SSL benchmark,
%% DiffusionDrive + SSL, and mechanistic analysis.
%% Ends with the Phase 0--4 experimental protocol.

% ============================================================================
\section{Exploratory Lightweight-Head Regime}
\label{sec:meth-lightweight-head}

The project began with an intentionally simple experimental regime: pair a pretrained visual backbone with a lightweight planning head and ask whether the representation alone already improves end-to-end driving behavior. The earliest version of this regime used a simple MLP waypoint regressor over image features and ego state. Its role was diagnostic rather than competitive. By minimizing planner capacity, it made it easier to attribute downstream changes to the backbone itself.

This regime remains important in the final thesis for two reasons. First, it explains why the thesis is fundamentally about representations rather than planner engineering. Second, it provides the historical starting point for the later pivot in scope. Once Drive-JEPA~\cite{wang-drive-jepa} demonstrated that V-JEPA pretraining plus a lightweight decoder could already be highly competitive on standard NAVSIM, the thesis moved away from asking whether SSL can improve in-distribution PDMS and toward the harder question of where that representational advantage remains uniquely informative.

% ============================================================================
\section{NAVSIM City Splits}
\label{sec:meth-city-splits}

\subsection{City Extraction from nuPlan Metadata}
\label{sec:meth-city-extraction}

City labels are recovered directly from the underlying nuPlan/OpenScene log metadata. Each scenario token inherits the city prefix of its source log, with prefixes such as \texttt{us-ma-boston}, \texttt{us-nv-las-vegas}, \texttt{us-pa-pittsburgh}, and \texttt{sg-one-north}. A preprocessing script walks the metadata once, assigns every scenario token to exactly one city, and writes city-specific token lists used by the training and evaluation dataloaders. This step is simple but essential: it turns NAVSIM from a pooled benchmark into a reproducible geographic-transfer benchmark whose train/test partitions are defined by city identity rather than random sampling.

\subsection{Split Statistics and Distribution}
\label{sec:meth-split-stats}

The cross-city protocol uses the natural geographic partitions already present in the source datasets. For nuScenes~\cite{caesar-nuscenes}, the Boston split contains 390 training scenes (2.12 hours) and 77 validation scenes (0.42 hours), while Singapore contains 310 training scenes (1.68 hours) and 73 validation scenes (0.40 hours). For NAVSIM~\cite{dauner-navsim}, the city distribution is more imbalanced: Las Vegas contributes 818 training scenes and 50 evaluation scenes, Boston 255 and 62, Pittsburgh 140 and 22, and Singapore 97 and 13. These counts are large enough to expose meaningful transfer behavior while still making the data imbalance itself part of the problem; Singapore is both the smallest domain and the only left-hand-traffic environment, which makes it a particularly stringent test of transfer~\cite{naeinian-cross-city-2026}. \reffig{fig:method-city-map} illustrates the four cities schematically alongside their train/test scene counts; the full data card is in \refapp{app:city-data-card}.

\begin{figure}[ht]
    \centering
    \begin{tikzpicture}[
        city/.style = {draw, rounded corners=3pt, minimum width=26mm, minimum height=14mm, align=center, font=\small},
        continent/.style = {draw, densely dashed, rounded corners=4pt, inner sep=4mm, font=\scriptsize\itshape, label position=above, label distance=-1mm},
        citycount/.style = {font=\scriptsize, gray},
    ]
        \node[city, fill=red!8] (lv) {Las Vegas (LV)\\{\scriptsize\color{gray}818 / 50}};
        \node[city, fill=red!8, right=10mm of lv] (pit) {Pittsburgh (P)\\{\scriptsize\color{gray}140 / 22}};
        \node[city, fill=red!8, right=10mm of pit] (bos) {Boston (B)\\{\scriptsize\color{gray}255 / 62}};
        \node[draw, densely dashed, rounded corners=4pt, inner sep=3mm, fit=(lv) (pit) (bos), label=above:\itshape North America, label distance=-1mm] (na) {};

        \node[city, fill=blue!8, right=18mm of bos] (sg) {Singapore (S)\\{\scriptsize\color{gray}97 / 13}\\\tiny left-hand traffic};
        \node[draw, densely dashed, rounded corners=4pt, inner sep=3mm, fit=(sg), label=above:\itshape Southeast Asia, label distance=-1mm] {};

        \node[font=\scriptsize, below=4mm of lv.south west, anchor=north west, text width=120mm] {\textit{trainval / test scene counts from \refapp{app:city-data-card}. Singapore is the only left-hand-traffic city and the smallest training source, making it a stringent cross-city target.}};
    \end{tikzpicture}
    \caption{Schematic of the four NAVSIM City Splits. Three North-American cities share traffic conventions (right-hand-side driving) and broadly comparable road-marking standards; Singapore adds a different traffic convention and distinct road-paint conventions. Scene counts are from \reftab{tab:city-splits}.}
    \label{fig:method-city-map}
\end{figure}

\subsection{Evaluation Protocols}
\label{sec:meth-eval-protocols}

The thesis uses a hierarchy of evaluation settings. The standard all-city protocol serves as a reference point: train and evaluate on the benchmark's original mixed-city split. The key stress test is strict single-city transfer, where a model is trained on one city and evaluated on unseen cities without fine-tuning, adaptation, or test-time modification. On nuScenes this yields a two-direction protocol (Boston$\rightarrow$Singapore and Singapore$\rightarrow$Boston); on NAVSIM it yields a full $4 \times 4$ train-city/test-city matrix. We also describe leave-one-city-out evaluation as a useful future extension, but the core empirical results in this thesis come from the first two settings because they match the controlled protocol introduced in~\cite{naeinian-cross-city-2026}.
\begin{enumerate}
    \item \textbf{All-city (standard)}: Train and test on the full NAVSIM split (all cities mixed). Reproduces published numbers.
    \item \textbf{Single-city (zero-shot transfer)}: Train on city A, test on all cities. Produces a $4 \times 4$ generalization matrix.
    \item \textbf{Leave-one-city-out (LOCO)}: Train on 3 cities, test on the held-out city. More realistic multi-city regime.
\end{enumerate}

For score-based metrics such as PDMS, we summarize out-of-distribution performance for a training city $c_i$ by averaging over the other cities~\cite{naeinian-cross-city-2026}:
\begin{equation}
    S_\text{OOD}(c_i)=\frac{1}{|C|-1}\sum_{c_j \neq c_i} S(c_i \rightarrow c_j).
\end{equation}

We additionally report the generalization gap as the difference between in-domain and out-of-domain closed-loop performance:
\begin{equation}
    \Delta_\text{gen} = \text{PDMS}_\text{in-dist} - \text{PDMS}_\text{OOD}.
\end{equation}

For open-loop error metrics, where lower is better, the more informative quantity is the error inflation ratio
\begin{equation}
    R_\text{err}(c_\text{train}\rightarrow c_\text{test})=\frac{E_\text{cross}(c_\text{train}\rightarrow c_\text{test})}{E_\text{in}(c_\text{train}\rightarrow c_\text{train})},
\end{equation}
which equals 1 for perfect preservation and grows as transfer degrades~\cite{naeinian-cross-city-2026}.

% ============================================================================
\section{SSL Pre-Training Protocol}
\label{sec:meth-ssl-pretraining}

To isolate the effect of the self-supervised objective from architecture and capacity, the Phase 1 protocol in~\cite{naeinian-cross-city-2026} performs domain-specific pretraining on nuScenes using the same ViT-S/14 backbone for I-JEPA, DINOv2-style distillation, and MAE. The goal is not to maximize absolute benchmark performance during pretraining, but to create matched representation initializations whose downstream behavior can be compared fairly under city transfer.

\begin{itemize}
    \item \textbf{Architecture}: ViT-S/14 for all domain-specific SSL runs
    \item \textbf{Pre-training data}: nuScenes training split (driving images)
    \item \textbf{Views and resolution}: all six camera views, with both square $224 \times 224$ and rectangular $224 \times 392$ inputs
    \item \textbf{Training}: 100 epochs with AdamW and cosine learning-rate scheduling
    \item \textbf{Hardware}: 2$\times$ H200 GPUs
    \item \textbf{Batch size}: 64 per GPU
\end{itemize}

Rectangular inputs preserve the native geometry of the driving scene, while square inputs align better with common ImageNet-style ViT pretraining. The comparison between those two settings turns out to matter empirically, particularly for I-JEPA under cross-city transfer~\cite{naeinian-cross-city-2026}. For generic large-scale pretraining, we use released ImageNet-scale checkpoints without additional driving-specific adaptation.

% ============================================================================
\section{SSL Backbone Integration}
\label{sec:meth-backbones}

\subsection{Backbone Adapter Interface}
\label{sec:meth-backbone-adapter}

All backbones are wrapped behind a common adapter interface so that planner code does not depend on model-specific preprocessing or tensor shapes. Conceptually, every backbone implements a common mapping
\[
\texttt{images} \mapsto \texttt{features},
\]
while the adapter handles input normalization, resizing, token pooling, and feature reshaping. This abstraction is important for the thesis because it makes backbone swapping a controlled intervention rather than a code fork. In practice, the same planning stack can be re-run with ResNet, I-JEPA, DINOv2, or MAE by changing configuration rather than rewriting the downstream model.

\subsection{Backbone Configurations}
\label{sec:meth-backbone-configs}

Table~\ref{tab:backbones} summarizes the backbone families used throughout the thesis. The key control principle is to vary the pretraining objective while keeping downstream planners fixed wherever possible.

\begin{table}[ht]
\centering
\caption{SSL backbone configurations.}
\label{tab:backbones}
\begin{tabular}{lllcl}
\toprule
\textbf{Backbone} & \textbf{SSL Type} & \textbf{Pre-train Data} & \textbf{Output Dim} & \textbf{Used In} \\
\midrule
ResNet-34         & Supervised      & ImageNet-1K & 512  & LAW, TF, DD \\
I-JEPA ViT-H/14  & Predictive      & ImageNet-1K & 1280 & LAW, TF \\
I-JEPA ViT-S/14  & Predictive      & nuScenes    & 384  & DD \\
DINOv2 ViT-L/14  & Discriminative  & LVD-142M    & 1024 & LAW, TF \\
MAE ViT-L/16     & Generative      & ImageNet-1K & 1024 & LAW, TF, DD \\
\bottomrule
\end{tabular}
\end{table}

TF denotes TransFuser and DD denotes DiffusionDrive. Unless stated otherwise, SSL backbones are frozen during downstream training. When fine-tuning is enabled for ablations, the backbone receives a smaller learning-rate multiplier than the planning head to reduce catastrophic drift away from the pretrained representation.

\subsection{Feature Projection}
\label{sec:meth-feature-proj}

Because the backbones emit features with different dimensionality and structure, a lightweight projection module standardizes them before they enter the planner. For global-regression heads, this is a linear projection from the backbone output dimension to the hidden dimension expected by the planner. For token-based backbones integrated into DiffusionDrive, the adapter projects ViT patch tokens $(B, N, C)$ into spatial feature maps $(B, C', H, W)$ compatible with the planner's multi-scale cross-attention blocks. This projection is intentionally lightweight so that the thesis remains focused on the representation rather than on heavy task-specific adaptation. \reffig{fig:method-adapter} sketches the data-shape flow.

\begin{figure}[ht]
    \centering
    \begin{tikzpicture}[
        node distance = 5mm and 7mm,
        box/.style = {draw, rounded corners=2pt, minimum height=8mm, minimum width=26mm, align=center, font=\small},
        shape/.style = {font=\scriptsize\ttfamily, gray},
        arrow/.style = {-{Stealth[length=2mm]}, thick},
    ]
        \node[box] (img) {Image\\$3 \times H_0 \times W_0$};
        \node[box, right=of img, fill=black!5] (vit) {ViT backbone\\(frozen)};
        \node[box, right=of vit] (tok) {Patch tokens\\$B \times N \times C$};
        \node[box, right=of tok] (rsh) {Reshape +\\linear proj.};
        \node[box, right=of rsh, fill=black!5] (map) {Spatial map\\$B \times C' \times H \times W$};
        \node[box, right=of map] (plan) {Planner\\(TF / LAW / DD)};
        \draw[arrow] (img) -- (vit);
        \draw[arrow] (vit) -- (tok);
        \draw[arrow] (tok) -- (rsh);
        \draw[arrow] (rsh) -- (map);
        \draw[arrow] (map) -- (plan);
    \end{tikzpicture}
    \caption{Backbone adapter schematic. A frozen ViT encoder emits $N$ patch tokens; a lightweight reshape and linear projection restructure them into a spatial feature map whose shape matches what the downstream planner expects. ResNet-34 backbones already emit a spatial map and therefore skip the reshape step. This adapter is the only module that differs between Phase-1 TransFuser/LAW and Phase-2 DiffusionDrive runs.}
    \label{fig:method-adapter}
\end{figure}

% ============================================================================
\section{Planning Architecture I: LAW with Swappable Backbones}
\label{sec:meth-law}

LAW (Latent World Model)~\cite{li-law} serves as the primary open-loop test bed for the cross-city SSL study. Following~\cite{naeinian-cross-city-2026}, the planner and training recipe are held fixed while only the backbone initialization changes. This isolates the representational contribution from downstream architectural confounders.

\subsection{Architecture Overview}
\label{sec:meth-law-arch}

The LAW experiments use the same downstream planner across all backbone variants. Its key components are:
\begin{itemize}
    \item Per-view spatial decoders (one per camera view)
    \item LID (Lift, Image, Depth) depth encoding for 3D-aware features
    \item Delta waypoint prediction (predicts offsets from previous waypoint)
    \item TransformerDecoder latent world model (auxiliary training loss)
\end{itemize}

The auxiliary world-model loss is useful here because it regularizes features toward future-state prediction without changing the core planning objective. Even so, the thesis treats LAW primarily as a controlled test bed: because the planner is fixed, differences in cross-city behavior can be attributed mainly to the backbone.

\subsection{Training Configuration}
\label{sec:meth-law-training}

For the nuScenes open-loop experiments, LAW is trained for 12 epochs with AdamW, base learning rate $5\times 10^{-5}$, weight decay 0.01, cosine annealing with minimum ratio $10^{-3}$, and gradient clipping at maximum L2 norm 35~\cite{naeinian-cross-city-2026}. Training uses mixed precision on a single NVIDIA L40S GPU with batch size 2. Frozen and fully fine-tuned backbone variants are both evaluated to distinguish representational quality from adaptation capacity.

\subsection{TransFuser Baseline}
\label{sec:meth-transfuser}

TransFuser~\cite{chitta-transfuser} provides the closed-loop NAVSIM baseline. Its camera encoder is replaced by SSL-pretrained backbones while the LiDAR branch is either retained (TransFuser) or removed in favor of a positional latent (Latent TransFuser), following the setup in~\cite{naeinian-cross-city-2026}. The NAVSIM runs use Adam with learning rate $10^{-4}$, zero weight decay, no warmup or scheduler, gradient clipping at norm 1.0, 20 epochs of training, and an effective batch size of 128 across four H200 GPUs. This pair of architectures is important because it tests whether the representation effect persists when LiDAR is removed.

% ============================================================================
\section{Planning Architecture II: DiffusionDrive + SSL}
\label{sec:meth-diffusiondrive}

DiffusionDrive serves as the generative-planning counterpart to LAW and TransFuser. It allows the thesis to test whether the representation-centric findings from the regression setting survive a substantial change in planning head. In this phase, SSL backbones are integrated into DiffusionDrive, and a camera-only variant is studied as an edge-deployment setting.

\subsection{DiffusionDrive Architecture}
\label{sec:meth-dd-arch}

The relevant ingredients of DiffusionDrive for this thesis are:
\begin{enumerate}
    \item \textbf{Anchor trajectories}: K-means clustering on training set trajectories produces 256 prototypical trajectories (\texttt{trajectory\_anchors\_256.npy}). These serve as the denoising starting point instead of Gaussian noise.
    \item \textbf{Truncated diffusion}: Only 2 denoising steps at inference (vs.\ 50--1000 in standard diffusion). Starts from nearest anchor trajectory, not pure noise. Enables 45+ FPS.
    \item \textbf{Multi-scale cross-attention decoder}: Conditions denoising on visual features at multiple spatial scales.
    \item \textbf{Original backbone}: ResNet-34 camera + ResNet-34 LiDAR (TransFuser-style fusion).
\end{enumerate}

\subsection{SSL Backbone Integration}
\label{sec:meth-dd-ssl}

Integrating ViT-style SSL encoders into DiffusionDrive requires a small set of architectural modifications:
\begin{enumerate}
    \item \textbf{Replace ResNet-34 image encoder} with I-JEPA ViT-S/14 (or MAE ViT-L/16). ViT outputs $(B, N, C)$ patch tokens.
    \item \textbf{Feature adapter}: Reshape ViT tokens to spatial feature maps $(B, C, H, W)$ via learned linear projection + reshape. A lightweight FPN (Feature Pyramid Network) generates multi-scale features compatible with DiffusionDrive's \texttt{MultiScaleCrossAttention}.
    \item \textbf{Two resolution modes}: Rectangular (native aspect, preserving horizontal FOV) and square (224$\times$224 center-crop, matching ImageNet pre-training). Results show resolution choice matters ($\sim$1pp impact).
    \item \textbf{Frozen vs.\ trainable}: Both modes tested. Frozen backbone + trainable adapter provides best balance.
\end{enumerate}

\subsection{Camera-Only Variant}
\label{sec:meth-dd-camera-only}

For the camera-only variant, all LiDAR-dependent inputs and rasterized BEV features are removed from the pipeline. The planner then operates only on camera images and ego state, with the missing LiDAR branch replaced by a lightweight learned latent input. This variant matters for two reasons: it targets lower-cost deployment settings, and it provides a stricter test of whether richer visual representations can compensate for missing geometric sensing.

\subsection{Training Configuration}
\label{sec:meth-dd-training}

The DiffusionDrive runs use AdamW with learning rate $10^{-4}$, weight decay $10^{-4}$, cosine learning-rate decay with a 3-epoch warmup, and distributed data parallel training. Most exploratory all-city experiments were shortened relative to the full recipe in order to compare more backbone variants under fixed compute; when this was done, the shortened runs were interpreted as architectural probes rather than as final leaderboard attempts.

% ============================================================================
\section{Experimental Protocol}
\label{sec:meth-exp-protocol}

Experiments are organized into one exploratory phase followed by four main phases, each building on the previous:

\subsection{Phase 0: SSL Backbone + Lightweight Head (Exploratory)}
\label{sec:meth-phase0}

Phase 0 captures the original hypothesis behind the thesis. The initial setup, first documented in~\cite{hamza-sanchez-navsim-report}, used a frozen I-JEPA encoder with a small MLP waypoint regressor over front-camera features and ego state. Later pilot runs expanded this to partial fine-tuning, three-camera fusion, and controlled TransFuser backbone swaps. The point of the phase was not to establish a final benchmark number, but to test whether the representation itself was already strong enough to matter when planner capacity was deliberately limited. In the final thesis this phase is interpreted qualitatively: it motivated the representation-first view of the problem, but ceased to be the central contribution once Drive-JEPA~\cite{wang-drive-jepa} established that JEPA-style pretraining could already be highly competitive on mixed-city NAVSIM.

\subsection{Phase 1: Cross-City SSL Benchmark}
\label{sec:meth-phase1}

Phase 1 aggregates the core benchmark first reported in~\cite{naeinian-cross-city-2026}. It combines open-loop LAW experiments on the two-city nuScenes split with closed-loop TransFuser and Latent TransFuser experiments on four-city NAVSIM. The backbone suite includes supervised baselines, ImageNet-pretrained SSL backbones, and domain-specific nuScenes-pretrained I-JEPA, DINOv2, and MAE variants under both frozen and fully fine-tuned downstream settings. The key output is not a single leaderboard score, but the change from in-domain to cross-city behavior.

\subsection{Phase 2: DiffusionDrive All-City}
\label{sec:meth-phase2}

Phase 2 revisits mixed-city NAVSIM after the appearance of Drive-JEPA. The purpose here is not simply to chase the highest in-distribution PDMS, but to test whether SSL benefits survive in a different generative planning architecture and in a camera-only regime. DiffusionDrive with SSL backbones on the standard NAVSIM split establishes this architectural bridge before any cross-city evaluation. The phase covers 9 configurations (Table~\ref{tab:diffusiondrive-allcity}).

\subsection{Phase 3: DiffusionDrive Cross-City}
\label{sec:meth-phase3}

Phase 3 extends the city-split protocol to DiffusionDrive. The intended comparison is single-city training with DiffusionDrive + I-JEPA versus DiffusionDrive + ResNet-34 across the four NAVSIM cities, yielding eight training runs and a full $4 \times 4$ transfer matrix for each backbone family. This phase is methodologically important even when partially complete, because it is the direct test of RQ5: whether the cross-city SSL advantage observed in regression planners persists when the planning head becomes generative.

\subsection{Phase 4: Mechanistic Analysis}
\label{sec:meth-phase4}

Phase 4 asks the question that remains after the earlier phases: why do SSL representations transfer better? This stage uses feature analysis (CKA, linear probing, attention maps, MMD, effective rank) without additional training, drawing on checkpoints from Phases 1--3. See Section~\ref{sec:exp-why-ssl} for results.

% ============================================================================
\section{Metrics}
\label{sec:meth-metrics}

For every experiment, we report:
\begin{itemize}
    \item \textbf{PDMS} (overall) and sub-scores: NC, DAC, EP, TTC, Comfort
    \item \textbf{EPDMS} (extended evaluation, where applicable --- DiffusionDrive models)
    \item \textbf{Generalization gap} $\Delta_\text{gen}$ and mean \textbf{OOD PDMS} $S_\text{OOD}$
    \item \textbf{Open-loop transfer inflation} $R_\text{err}$ for L2 displacement and collision rate, as defined in~\cite{naeinian-cross-city-2026}
    \item \textbf{Inference speed} (FPS) for DiffusionDrive variants
    \item All results tracked via Weights \& Biases
\end{itemize}

% ============================================================================
\section{Implementation Details}
\label{sec:meth-implementation}

Experiments were implemented in PyTorch and PyTorch Lightning, with Hydra-style configuration used to manage backbone, dataset, and optimizer variants. Training was run on the NYU Greene HPC cluster using NVIDIA L40S and H200 GPUs, with distributed data parallelism where required by model size and batch size. Pretrained checkpoints for public SSL backbones were loaded from their released model weights and wrapped through the common adapter interface described above. Experiment orchestration, checkpoint tracking, and evaluation were handled through the local thesis codebase and associated command-line management scripts.
