%!TEX root = ../thesis.tex
\chapter{Related Work}
\label{chap:relatedwork}

%% ~8 pages. Organized around three threads, not chronologically.
%% Each section ends with a gap statement leading to our contributions.

% ============================================================================
\section{Self-Supervised Learning for Autonomous Driving}
\label{sec:rw-ssl-driving}

Self-supervised learning has become a compelling alternative to supervised ImageNet pretraining for autonomous driving because it can exploit the scale of unlabelled driving logs while learning invariances that are closer to the downstream planning problem. The three SSL families most relevant to this thesis are predictive JEPA-style models, self-distillation methods such as DINOv2, and generative masked-image models such as MAE. Naeinian et al.~\cite{naeinian-cross-city-2026} provide the first controlled comparison of these paradigms under explicit geographic transfer for end-to-end planning, rather than only perception.

\subsection{Contrastive and Self-Distillation Methods}
Contrastive and self-distillation methods have been widely adopted for driving perception, where they improve data efficiency for detection and segmentation but are usually evaluated in perception-only settings. DINOv2 is especially attractive as a generic visual backbone because of its strong transfer across downstream tasks, yet prior to~\cite{naeinian-cross-city-2026} it had not been systematically tested for zero-shot cross-city planning robustness.

\subsection{Masked Image Modeling}
Masked image modeling offers a different trade-off. MAE learns from aggressive patch masking and pixel-space reconstruction, which can yield strong generic features while remaining architecturally simple~\cite{he-mae}. In autonomous driving, however, its use has been more limited on the planning side than on perception benchmarks. The cross-city study of~\cite{naeinian-cross-city-2026} is therefore important because it shows that MAE can remain competitive under geographic shift, especially when domain-specifically pretrained on nuScenes and transferred to the closed-loop NAVSIM setting.

\subsection{Joint-Embedding Predictive Architectures}
JEPA-style methods are particularly relevant because they predict latent semantics rather than reconstructing pixels, which in principle should reduce sensitivity to superficial appearance cues. I-JEPA provides the image-based formulation used throughout this thesis~\cite{assran-ijepa}, while V-JEPA extends the same intuition to video. In driving, the JEPA direction has quickly become important:
\begin{itemize}
    \item \textbf{AD-L-JEPA} applies latent JEPA pretraining in the driving domain, but focuses on perception rather than end-to-end planning.
    \item \textbf{Drive-JEPA} couples a V-JEPA-style backbone with proposal-centric planning and simulator trajectory distillation, reaching 93.3 PDMS on NAVSIM v1 and 87.8 EPDMS on NAVSIM v2 under the standard mixed-city protocol~\cite{wang-drive-jepa}. Its perception-free setting is equally notable: a V-JEPA encoder with a simple transformer decoder reaches 89.0 PDMS, showing that strong video pretraining can already make lightweight planning heads highly competitive.
\end{itemize}

\textbf{Gap}. For this thesis, Drive-JEPA is important not only as related work but as an inflection point. Once JEPA-style methods had already demonstrated strong mixed-city benchmark performance, the unresolved question was no longer whether SSL can improve PDMS at all. The more interesting gap became whether SSL offers a distinctive advantage under geographic shift, how that advantage interacts with planning-head design, and what feature-space mechanism explains it. The closest work to that question is Naeinian et al.~\cite{naeinian-cross-city-2026}, which isolates strict city-disjoint transfer on nuScenes and NAVSIM. The remaining gaps addressed in this thesis are broader mechanistic analysis, diffusion-based planning, and camera-only deployment.

\subsection{Pre-Training Data: ImageNet vs.\ Driving-Specific}
An important design choice is whether to rely on generic large-scale pretraining or on smaller but domain-aligned driving corpora. The evidence in~\cite{naeinian-cross-city-2026} suggests that this distinction matters more under transfer than it does on mixed-city validation. Large ImageNet-pretrained backbones can remain competitive in-domain, but domain-specific nuScenes pretraining substantially narrows the Boston$\rightarrow$Singapore generalization gap and produces the most stable closed-loop gains on NAVSIM. This thesis uses both regimes because they probe different sources of robustness: scale-induced semantic richness versus driving-specific invariances.

% ============================================================================
\section{Cross-Domain Generalization in Driving}
\label{sec:rw-cross-domain}

\subsection{Weather and Lighting Robustness}
Most domain-generalization work in autonomous driving focuses on weather, lighting, or simulation-to-real transfer. Those settings are important, but they primarily perturb appearance. Geographic transfer is harder because it also changes road topology, map priors, lane semantics, and traffic convention. The survey and experiments in~\cite{naeinian-cross-city-2026} argue that these structural shifts are precisely where conventional end-to-end benchmarks are weakest.

\subsection{Geographic / Cross-City Transfer}
Cross-city transfer has been studied only sparsely, and most of that work is still perception-centric. The recent zero-shot protocol of~\cite{naeinian-cross-city-2026} shows why planning requires separate analysis: the direction of transfer matters. Boston$\rightarrow$Singapore is much harder than Singapore$\rightarrow$Boston in open-loop LAW experiments, which indicates that geographic generalization is not symmetric and cannot be captured by a single pooled validation score. This directional effect is especially pronounced when transferring between right-hand and left-hand driving environments.
\begin{itemize}
    \item Most prior work focuses on perception metrics such as mAP rather than planning metrics such as PDMS.
    \item Zero-shot geographic transfer for end-to-end planning remained effectively unbenchmarked until the protocol introduced in~\cite{naeinian-cross-city-2026}.
    \item NAVSIM originally exposes aggregate splits, so per-city analysis required an explicit benchmark construction step.
\end{itemize}

\textbf{Gap}. The resulting benchmark gap is central to this thesis: once per-city NAVSIM evaluation is available, representation quality becomes visible as the dominant factor behind robustness.

\subsection{Domain Generalization Techniques}
Classical domain generalization techniques in driving modify the training procedure through augmentation, domain randomization, meta-learning, or invariance objectives. The approach taken in~\cite{naeinian-cross-city-2026} and extended in this thesis is deliberately simpler: keep the downstream planner and optimization recipe fixed, and change only the representation initialization. The empirical result is striking because this single intervention already collapses much of the cross-city gap, which suggests that the dominant failure mode lies in the backbone rather than in planner-specific training heuristics.

% ============================================================================
\section{Diffusion-Based Trajectory Planning}
\label{sec:rw-diffusion-planning}

\subsection{Full Diffusion Planners}
Full diffusion planners model multi-modal trajectory distributions by iteratively denoising from noise. Their main strength is expressive trajectory generation; their main weakness is cost. Running dozens or hundreds of denoising steps is often too slow for realistic real-time driving stacks, which is why this line of work is less directly useful for the thesis than accelerated variants.

\subsection{Truncated and Accelerated Diffusion}
DiffusionDrive is the main accelerated-diffusion reference point for this thesis. Instead of starting from pure Gaussian noise, it initializes from clustered anchor trajectories and uses only a small number of denoising steps. This preserves much of the flexibility of diffusion planning while keeping inference practical. That design makes it ideal for the thesis: it is architecturally different enough from LAW and TransFuser to test whether the SSL effect survives a head change, but efficient enough to run at the scale required for backbone ablations.

Follow-up work adds RL-style post-training to diffusion planners, improving in-distribution driving quality further. For the purposes of this thesis, these methods matter less as direct baselines than as evidence that planning-head sophistication is advancing rapidly. This strengthens the argument for focusing the thesis on the part of the pipeline that remains under-explained across architectures: the visual representation.

\subsection{Camera-Only Planners}
Camera-only planning is especially relevant because it removes the geometric crutch of LiDAR and therefore makes representation quality more visible. Methods such as PRIX and GoalFlow show that camera-only generative planning can be competitive, but the literature still leaves two gaps open for this thesis: camera-only diffusion planners have usually relied on supervised backbones, and diffusion planners have not been systematically evaluated for cross-city transfer. The JEPA-Diffusion phase of the thesis is positioned directly in that gap.

% ============================================================================
\section{Reinforcement Learning for Planning Refinement}
\label{sec:rw-rl-planning}

RL post-training for driving planners is an emerging direction rather than a settled recipe. The common idea is to refine a learned planner with a simulator-derived reward that better reflects downstream driving quality than pure imitation loss.
\begin{itemize}
    \item Diffusion-style planners can be improved with reward-guided post-training that uses simulator scoring rather than only imitation targets.
    \item Preference-based methods such as DPO offer an alternative in which trajectories are ranked pairwise rather than optimized through an explicit scalar reward.
    \item Classical policy-gradient fine-tuning remains relevant in closed-loop simulators such as CARLA, though it is usually more expensive and harder to stabilize.
\end{itemize}

\textbf{Gap}. Existing RL-style refinements are almost always reported on the same mixed distribution used for training. Whether those gains survive transfer to unseen cities is unknown. This motivates RQ6, which asks whether policy-improvement gains are genuinely general or merely city-specific.

% ============================================================================
\section{Summary and Positioning}
\label{sec:rw-summary}

The literature now supports three claims. First, SSL backbones are effective for end-to-end planning in standard mixed-city evaluation. Second, JEPA-style pretraining is especially promising at scale, as shown most clearly by Drive-JEPA~\cite{wang-drive-jepa}. Third, once explicit geographic splits are introduced, the choice of representation becomes a first-order robustness variable~\cite{naeinian-cross-city-2026}. This thesis is positioned at the intersection of those claims. It begins from the same lightweight-head intuition that motivated early JEPA experiments, acknowledges that Drive-JEPA largely settled the in-distribution performance question, and then moves to the more differentiated problem of cross-city robustness and its mechanism.

Gaps this thesis fills:
\begin{enumerate}
    \item No cross-city generalization benchmark exists for NAVSIM $\to$ \textbf{our city splits (Ch.~\ref{chap:methodology})}, first instantiated in~\cite{naeinian-cross-city-2026}.
    \item No systematic comparison of SSL backbones for planning generalization had been available before~\cite{naeinian-cross-city-2026} $\to$ \textbf{our thesis integrates and extends that 19-configuration study (Ch.~\ref{chap:experiments}, Phase 1)}.
    \item No camera-only planner combines SSL + truncated diffusion + EPDMS $\to$ \textbf{JEPA-Diffusion (Ch.~\ref{chap:experiments}, Phase 2)}.
    \item No mechanistic analysis of \emph{why} SSL improves driving generalization $\to$ \textbf{CKA/linear probing/MMD analysis (Ch.~\ref{chap:experiments}, Phase 4)}.
    \item No study tests whether RL post-training gains transfer across cities $\to$ \textbf{RQ6 (Ch.~\ref{chap:experiments}, exploratory)}.
\end{enumerate}
