%!TEX root = ../thesis.tex
\chapter{Experiments and Results}
\label{chap:experiments}

%% This chapter presents all experimental results.
%% Structure: 4 phases matching the methodology, plus "Why SSL Works" analysis.
%% Approx. 20 pages — the core of the thesis.

% ============================================================================
\section{Baseline Validation}
\label{sec:exp-baselines}

\textit{Validate that our implementations reproduce published numbers on the standard (non-city-split) NAVSIM benchmark before running cross-city experiments.}

% TODO: Expand into a proper table with sub-metrics
\begin{table}[h]
\centering
\caption{Baseline validation on NAVSIM (all cities, standard split).}
\label{tab:baselines}
\begin{tabular}{llc}
\toprule
\textbf{Exp} & \textbf{Configuration} & \textbf{PDMS} \\
\midrule
A3  & TransFuser (ResNet34, C+L) & 84.0 \\
A10 & DiffusionDrive (ResNet34, C+L) & 88.1 \\
\bottomrule
\end{tabular}
\end{table}

\textit{Both baselines match published results, validating our framework and training pipeline. These serve as reference points for all subsequent experiments.}

% ============================================================================
\section{Phase 1: Cross-City SSL Benchmark (IROS Results)}
\label{sec:exp-phase1}

\textit{This section presents the core findings from our IROS 2026 submission: a systematic comparison of SSL backbones under zero-shot cross-city transfer using TransFuser and LAW architectures.}

\subsection{Experimental Setup}
\textit{19 backbone configurations (Table I from IROS) evaluated on nuScenes with per-city splits. Backbones: ResNet34 (supervised), I-JEPA ViT-H/14, DINOv2 ViT-L/14, MAE ViT-L/16. Both frozen and fine-tuned variants. Two planning heads: TransFuser and LAW.}

\subsection{In-Distribution Performance}
\label{sec:exp-phase1-indist}

\textit{Import IROS Table I: Full nuScenes results across all 19 backbone configurations. Present in-distribution PDMS for each backbone $\times$ city combination. Key observation: SSL backbones achieve comparable or slightly lower in-distribution PDMS vs.\ supervised ResNet ($\sim$1pp gap).}

\subsection{Cross-City Transfer Matrices}
\label{sec:exp-phase1-matrix}

\textit{Import IROS Fig 3: $4 \times 4$ train-city $\times$ test-city matrices for each backbone. Heatmap visualization showing transfer asymmetry. Key findings:}
\begin{itemize}
    \item \textit{Which backbone has the smallest generalization gap?}
    \item \textit{Which city is the best training ground? (RQ2)}
    \item \textit{Transfer asymmetry: city A $\to$ B $\neq$ city B $\to$ A}
\end{itemize}

\subsection{OOD PDMS Analysis}
\label{sec:exp-phase1-ood}

\textit{Import IROS Fig 4: 8-panel NAVSIM OOD PDMS figure. Transfer ratio analysis. Core finding: SSL reduces cross-city generalization gaps by an order of magnitude. State the central thesis claim with concrete numbers:}

\begin{quote}
\textit{``Supervised ResNet34 shows a transfer ratio inflation of X\% under cross-city evaluation, while I-JEPA reduces this to Y\%, demonstrating that cross-city generalization failure is fundamentally a representation problem, not a planning architecture problem.''}
\end{quote}

\subsection{Transfer Asymmetry Analysis}
\label{sec:exp-phase1-asymmetry}

\textit{Analyze directional transfer patterns. Some city pairs exhibit strong asymmetry (training on A transfers well to B, but not vice versa). Relate to visual/structural properties of each city (road density, vegetation, infrastructure style).}

\subsection{Per-Metric Breakdown}
\label{sec:exp-phase1-metrics}

\textit{Break down PDMS into sub-metrics (NC, DAC, EP, TTC, Comfort). Which sub-metrics suffer most under cross-city transfer? Do SSL backbones uniformly improve all sub-metrics or only certain ones (e.g., drivable area compliance may be more city-dependent than collision avoidance)?}

% ============================================================================
\section{Phase 2: DiffusionDrive All-City Results}
\label{sec:exp-phase2}

\textit{Present the full set of DiffusionDrive experiments on the standard all-city NAVSIM split. This establishes whether SSL backbones transfer their benefits from regression-based planners (LAW/TransFuser) to generative planners (DiffusionDrive).}

\begin{table}[h]
\centering
\caption{DiffusionDrive + SSL backbone results on NAVSIM (all cities).}
\label{tab:diffusiondrive-allcity}
\begin{tabular}{lllcc}
\toprule
\textbf{Exp} & \textbf{Backbone} & \textbf{Modality} & \textbf{Frozen} & \textbf{PDMS} \\
\midrule
A12   & ResNet34              & C+L & ---       & 87.9 \\
A12-o & MAE (rect, trainable) & C+L & No        & 86.9 \\
A12-u & MAE (sq, trainable)   & C+L & No        & 86.8 \\
A12-q & I-JEPA (sq, trainable)& C+L & No        & 86.2 \\
A10-j & I-JEPA nuSc (rect, frozen) & C+L & Yes  & 85.2 \\
A10-p & I-JEPA nuSc (sq, frozen)   & C+L & Yes  & 86.3 \\
A13   & ResNet34 (camera-only)& C   & ---       & 86.6 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{SSL vs.\ Supervised in Generative Planning}
\label{sec:exp-phase2-ssl-vs-sup}

\textit{The $\sim$1pp gap between SSL (86.2--86.9) and supervised ResNet34 (87.9) on all-city training mirrors the pattern from Phase 1. SSL features are slightly weaker in-distribution. However, this gap is the setup for the cross-city punchline in Phase 3: SSL may cost 1\% in-distribution but save 10\%+ out-of-distribution.}

\subsection{Camera-Only: SSL Partially Replaces LiDAR}
\label{sec:exp-phase2-camera-only}

\textit{ResNet34 camera-only (A13) achieves 86.6 PDMS, only 1.3pp below C+L (87.9). This is a remarkably small gap. Present the hypothesis: SSL features, by encoding richer spatial structure, may further close this gap. The I-JEPA camera-only experiment (pending) will test this directly.}

% TODO: Add I-JEPA camera-only result when available

\subsection{Frozen vs.\ Fine-Tuned Backbone}
\label{sec:exp-phase2-frozen}

\textit{Compare A10-j (frozen, 85.2) vs.\ A10-p (frozen sq, 86.3) vs.\ A12-q (trainable, 86.2). Resolution matters more than fine-tuning for I-JEPA in DiffusionDrive. Frozen square-crop (86.3) matches trainable (86.2), suggesting the frozen backbone acts as implicit regularization without sacrificing performance. This has implications for cross-city transfer.}

% ============================================================================
\section{Phase 3: DiffusionDrive Cross-City Protocol}
\label{sec:exp-phase3}

\textit{\textbf{CRITICAL PENDING EXPERIMENT.} This is the experiment that completes the thesis narrative. Train DiffusionDrive + I-JEPA and DiffusionDrive + ResNet34 on each single city, evaluate on all 4 cities. If the IROS pattern holds (SSL reduces transfer ratio inflation), the thesis claim extends from regression planners to generative planners.}

\subsection{Experimental Setup}
\textit{4 single-city training runs $\times$ 2 backbones (I-JEPA nuSc sq trainable, ResNet34) = 8 training runs. Each evaluated on all 4 city test sets = 32 evaluation jobs. Use best all-city configs from Phase 2.}

\subsection{Cross-City Generalization Matrices}
\textit{Present $4 \times 4$ matrices for both backbones. Compare transfer ratios with Phase 1 (LAW/TransFuser) results.}

% TODO: Fill with results once experiments complete

\subsection{Regression vs.\ Diffusion Head Under Domain Shift}
\label{sec:exp-phase3-head-comparison}

\textit{Direct comparison: does the diffusion planning head independently affect cross-city generalization? Compare:}
\begin{itemize}
    \item I-JEPA + LAW regression head (Phase 1) vs.\ I-JEPA + DiffusionDrive (this phase)
    \item ResNet34 + LAW (Phase 1) vs.\ ResNet34 + DiffusionDrive (this phase)
\end{itemize}
\textit{If SSL improvements are consistent across both planning heads, this confirms the thesis claim that representation quality is the primary factor. (RQ5)}

\subsection{The ``1\% In-Distribution, 10\% OOD'' Narrative}
\label{sec:exp-phase3-narrative}

\textit{Synthesize the key strategic insight: SSL costs $\sim$1pp in all-city/in-distribution PDMS but saves an order of magnitude in cross-city generalization gap. Present this as the central practical recommendation of the thesis.}

% ============================================================================
\section{Why Does SSL Improve Driving Planning?}
\label{sec:exp-why-ssl}

\textit{This section elevates the thesis from ``we show SSL helps'' to ``we understand \textbf{why} SSL helps.'' The analyses below require no additional training --- only feature extraction and linear algebra on existing checkpoints.}

\subsection{Representation Similarity Across Cities (CKA)}
\label{sec:exp-cka}

\textit{Compute Centered Kernel Alignment (CKA) between feature representations extracted from Boston vs.\ Singapore images, comparing SSL vs.\ supervised backbones.}

\textit{\textbf{Hypothesis}: SSL features from different cities are more similar (higher CKA) than supervised features, meaning SSL learns city-invariant representations.}

\textit{\textbf{Method}: Extract penultimate-layer features from 1000 images per city. Compute linear CKA for all city pairs. Compare ResNet34 vs.\ I-JEPA vs.\ MAE vs.\ DINOv2. Present as a heatmap.}

% TODO: CKA heatmap figure

\subsection{Linear Probing for City Classification}
\label{sec:exp-linear-probe}

\textit{Train a linear classifier on frozen backbone features to predict city identity.}

\textit{\textbf{Hypothesis}: Supervised features are more city-discriminative (higher accuracy) than SSL features. If SSL features cannot distinguish cities well, they don't encode city-specific information, explaining why they transfer better.}

\textit{\textbf{Method}: Extract features from all training images, fit logistic regression, report classification accuracy. If ResNet34 achieves 90\%+ city classification and I-JEPA achieves $\sim$60\%, that directly explains the generalization gap.}

% TODO: City classification accuracy table

\subsection{Feature Visualization via Attention Maps}
\label{sec:exp-attention}

\textit{For ViT-based SSL backbones, visualize CLS token attention maps on the same driving scene from different cities.}

\textit{\textbf{Hypothesis}: SSL attention focuses on road structure (lanes, intersections, vehicles) while supervised features correlate with city-specific textures (vegetation, building styles, road surface).}

\textit{\textbf{Method}: Pick 6--8 representative scenes per city, extract last-layer CLS attention maps, present side-by-side comparisons. Qualitative but visually compelling evidence for the mechanism.}

% TODO: Attention map figure (2x4 grid: SSL vs supervised × 4 cities)

\subsection{Feature Distribution Shift (MMD)}
\label{sec:exp-mmd}

\textit{Compute Maximum Mean Discrepancy (MMD) between feature distributions of train-city vs.\ test-city for each backbone.}

\textit{\textbf{Hypothesis}: SSL backbones show smaller MMD (less distribution shift in feature space), confirming they produce more domain-invariant representations.}

\textit{\textbf{Method}: Extract features, compute RBF-kernel MMD for each (train-city, test-city, backbone) triple. Present as grouped bar chart.}

% TODO: MMD bar chart figure

\subsection{Frozen Backbone as Implicit Regularization}
\label{sec:exp-implicit-reg}

\textit{Compare the effective rank of learned planning head weights when trained on frozen SSL vs.\ supervised features.}

\textit{\textbf{Hypothesis}: Frozen SSL encoders constrain the planning head to operate on a more structured feature manifold, acting as implicit regularization. The planning head learns a lower-rank mapping, which generalizes better.}

\textit{\textbf{Method}: Compute singular value spectrum of planning head weight matrices for each backbone configuration. Compare effective rank (number of significant singular values) and condition number.}

% TODO: Singular value spectrum figure

\subsection{Synthesis: A Mechanistic Account}
\label{sec:exp-why-synthesis}

\textit{Combine findings from all analyses into a coherent mechanistic explanation:}
\begin{enumerate}
    \item \textit{SSL pre-training produces features that are less city-discriminative (linear probing) and more similar across cities (CKA).}
    \item \textit{SSL attention focuses on task-relevant structure (road layout, vehicles) rather than city-specific textures (attention maps).}
    \item \textit{The resulting feature space has smaller distributional shift under city transfer (MMD).}
    \item \textit{Frozen SSL backbones implicitly regularize the planning head, preventing overfitting to city-specific patterns (effective rank).}
\end{enumerate}

% ============================================================================
\section{Ablation Studies}
\label{sec:exp-ablations}

\subsection{Input Resolution: Rectangular vs.\ Square Crop}
\textit{Compare rectangular (native aspect) vs.\ square (center-crop) input preprocessing for ViT backbones. Phase 2 results show this matters: I-JEPA sq frozen (86.3) vs.\ rect frozen (85.2). Discuss why square crops may preserve more spatial information for ViTs trained on square ImageNet images.}

\subsection{Frozen vs.\ Fine-Tuned Backbone (Extended)}
\textit{Extend the Phase 2 frozen/fine-tuned comparison with cross-city evaluation. \textbf{Hypothesis}: Fine-tuning improves in-distribution but hurts cross-city transfer. If confirmed, this supports the implicit regularization narrative from Sec.~\ref{sec:exp-implicit-reg}.}

\subsection{Pre-Training Data: ImageNet vs.\ nuScenes}
\textit{Compare I-JEPA pre-trained on ImageNet-1K vs.\ I-JEPA pre-trained on nuScenes driving data. Does domain-specific pre-training improve cross-city transfer, or does ImageNet's diversity suffice?}

\subsection{ViT Scale: ViT-S vs.\ ViT-H}
\textit{Compare I-JEPA ViT-S/14 (used in DiffusionDrive for speed) vs.\ ViT-H/14 (used in cross-city LAW study). Trade-off: accuracy vs.\ inference cost. Does scale affect generalization proportionally?}

% ============================================================================
\section{EPDMS Evaluation}
\label{sec:exp-epdms}

\textit{Report Extended PDM Scores on best DiffusionDrive models. Even if not SOTA, reporting EPDMS for a camera-only SSL method is a contribution --- no prior work does this.}

% TODO: EPDMS table when available

\textit{Compare with published EPDMS: PRIX (84.2), Drive-JEPA (87.8), DiffusionDriveV2 (85.5). Contextualize our camera-only result.}

% ============================================================================
\section{RL Post-Training Transfer (Exploratory)}
\label{sec:exp-rl}

\textit{If conducted: apply GRPO/DPO to best all-city JEPA-Diffusion model. Measure PDMS improvement, then re-evaluate on held-out cities. (RQ6)}

\textit{The finding is valuable either way:}
\begin{itemize}
    \item \textit{If RL gains transfer: SSL + RL is a viable recipe for robust, high-performance planning.}
    \item \textit{If RL gains don't transfer: RL overfits to city-specific reward landscapes, an important negative result.}
\end{itemize}

% ============================================================================
\section{Qualitative Analysis}
\label{sec:exp-qualitative}

\textit{Visualize predicted trajectories for selected scenarios:}
\begin{itemize}
    \item \textit{Success cases: where SSL backbone succeeds and supervised fails under cross-city transfer.}
    \item \textit{Failure cases: where all backbones struggle (e.g., unusual intersection geometries, rare traffic patterns).}
    \item \textit{Diffusion trajectory diversity: show multi-modal trajectory samples from DiffusionDrive, comparing SSL vs.\ supervised anchor selection.}
\end{itemize}

% ============================================================================
\section{Discussion}
\label{sec:exp-discussion}

\textit{Synthesize findings across all phases into the thesis argument:}

\begin{enumerate}
    \item \textbf{Cross-city failure is a representation problem.} Confirmed by Phase 1 (regression planners) and Phase 3 (generative planners). SSL reduces generalization gaps consistently across planning architectures.

    \item \textbf{The planning head architecture has secondary impact.} Compare Phase 1 vs.\ Phase 3 transfer ratios. If similar, the backbone dominates generalization. If different, discuss what the diffusion head adds.

    \item \textbf{SSL features are city-invariant by construction.} The ``Why SSL Works'' analyses (CKA, linear probing, MMD) provide a mechanistic explanation, not just empirical observation.

    \item \textbf{The 1\% in-distribution cost is worth the 10\%+ OOD gain.} Practical deployment recommendation.

    \item \textbf{Limitations and threats to validity:}
    \begin{itemize}
        \item Open-loop evaluation only (PDMS/EPDMS); closed-loop would strengthen claims.
        \item Limited to 4 cities; broader geographic diversity needed.
        \item SSL backbones pre-trained on ImageNet, not driving data.
        \item DiffusionDrive cross-city results still pending at time of writing (if applicable).
    \end{itemize}
\end{enumerate}
