%!TEX root = ../thesis.tex
\chapter{Conclusions and Future Work}
\label{chap:conclusions}

%% ~4 pages. Draft from IROS conclusion, expand with DiffusionDrive findings.

% ============================================================================
\section{Summary of Contributions}
\label{sec:concl-summary}

This thesis evolved through three linked questions rather than a single benchmark objective. It began from the hypothesis that SSL backbones should improve even lightweight end-to-end planners. The publication of Drive-JEPA~\cite{wang-drive-jepa} confirmed that JEPA-style pretraining could indeed deliver very strong mixed-city performance, which pushed the thesis toward a sharper and more differentiated question: what does SSL buy beyond standard leaderboard PDMS? The answer developed in this thesis is cross-city robustness and the representational mechanism behind it.

We summarize the resulting contributions:

\begin{enumerate}
    \item \textbf{NAVSIM City Splits}: We created the first per-city benchmark for NAVSIM, partitioning scenarios across Boston, Pittsburgh, Las Vegas, and Singapore. This benchmark enables systematic evaluation of cross-city transfer that was previously impossible on NAVSIM, and the first phase of that analysis was reported in~\cite{naeinian-cross-city-2026}.

    \item \textbf{Cross-City SSL Generalization}: Through 19 backbone configurations across TransFuser and LAW architectures, we demonstrated that self-supervised visual representations (I-JEPA, DINOv2, MAE) reduce cross-city generalization gaps by an order of magnitude compared to supervised baselines. In the most extreme open-loop setting reported in~\cite{naeinian-cross-city-2026}, Boston$\rightarrow$Singapore transfer falls from 9.77$\times$ L2 inflation and 19.34$\times$ collision inflation under a supervised backbone to 1.20$\times$ and 0.75$\times$ under domain-specific I-JEPA. This is the central finding of the thesis: \textbf{cross-city failure is a representation problem, not a planning architecture problem}.

    \item \textbf{DiffusionDrive + SSL}: We integrated SSL backbones into DiffusionDrive, achieving up to 86.9 PDMS (MAE, C+L) and demonstrating a camera-only variant at 86.6 PDMS (ResNet-34). The direct test of whether the representation-centric finding survives the substitution of a generative planner for a regression one is specified as Phase~3 (\refsec{sec:exp-phase3}) and scheduled as a follow-up: eight single-city training runs $\times$ 32 cross-city evaluations, with committed quantitative predictions (15--20 PDMS drop for supervised, 8--12 for frozen SSL, preserving the Phase-1 2$\times$ ratio). The I-JEPA camera-only comparison is similarly specified (\refsec{sec:exp-phase2-camera-only}) and awaits a single training run.

    \item \textbf{Mechanistic Analysis}: Our ``Why does SSL work?'' investigation covers 188 analyzed checkpoints --- linear probing, CKA, MMD, and feature-covariance effective rank --- and produces three calibrated findings. (a) Supervised ResNet-34 encodes city identity near-losslessly (linear probe 0.97--0.98), while frozen SSL encoders score 0.74--0.90; in matched Boston-only comparisons the probe-accuracy ordering mirrors the OOD PDMS drop ordering exactly (ResNet-34 0.97 / drop 16.1, MAE 0.89 / 13.2, I-JEPA 0.83 / 9.6, DINOv2 0.81 / 6.3). (b) Pretraining objective matters beyond this probe signal: MAE frozen LatTF is the best family in the roster (OOD drop 6.9, total PDMS 64.2) despite moderately high probe accuracy, while DINOv2 most frequently collapses under fine-tune (three runs with feature-covariance effective rank $<5$, including A7-b at rank 1.2 and A6-b I-JEPA at rank 2.9). (c) The operational lever is freezing: fine-tuning compresses feature-covariance rank for every SSL family (e.g., MAE 72 $\to$ 47), eroding the pretrained invariances on which the transfer result depends.
\end{enumerate}

% ============================================================================
\section{Key Findings}
\label{sec:concl-findings}

The project-level takeaway is therefore broader than any one phase: lightweight-head experiments motivated the representation hypothesis, Drive-JEPA clarified that SSL can already be powerful in-distribution, and the final thesis identifies zero-shot cross-city transfer as the regime where SSL's distinctive value becomes clearest.

Actionable takeaways for practitioners:

\begin{enumerate}
    \item \textbf{SSL pre-training is the single most impactful change for cross-city robustness.} Swapping a supervised backbone for a frozen SSL ViT, with no planner redesign, can reduce open-loop transfer inflation by nearly an order of magnitude and improve closed-loop OOD PDMS by several points~\cite{naeinian-cross-city-2026}.

    \item \textbf{The in-distribution cost is small; the OOD gain is large.} SSL backbones achieve $\sim$1pp lower PDMS in-distribution across every Phase-2 DiffusionDrive configuration we measured (SSL 86.2--86.9 vs ResNet-34 87.9, \reftab{tab:diffusiondrive-allcity}), but maintain performance under cross-city transfer where supervised baselines collapse: the Phase-1 Boston-only Latent~TransFuser shift from ResNet-34 to frozen I-JEPA reduces OOD PDMS drop from 16.1 to 9.6 in exchange for a 4-point in-distribution cost (\reftab{tab:probe-boston}). Phase~3 (\refsec{sec:exp-phase3}) will test whether this trade-off survives the planner swap to DiffusionDrive.

    \item \textbf{Frozen backbones generalize better than fine-tuned.} Fine-tuning improves in-distribution performance but may overfit to city-specific features. Frozen SSL backbones act as implicit regularization.

    \item \textbf{Camera-only planning is viable with SSL features.} ResNet-34 camera-only in DiffusionDrive loses only 1.3 PDMS relative to its camera+LiDAR variant (86.6 vs 87.9 PDMS), a narrower gap than the conventional view of LiDAR-as-indispensable would predict. The I-JEPA camera-only experiment outlined in \refsec{sec:exp-phase2-camera-only} is designed to test whether frozen SSL features close the residual gap entirely; the thesis commits to reporting the result as a defense-slide addendum rather than restructuring the argument around it.

    \item \textbf{The planning head has secondary impact on generalization.} Phase-1 LAW and TransFuser results both show the same representation-centric pattern despite architectural differences, and the Phase-2 DiffusionDrive results replicate the $\sim$1\,pp SSL in-distribution cost. Whether the pattern survives the substitution of a generative planner for regression under \emph{cross-city} shift is the direct empirical content of Phase~3 (\refsec{sec:exp-phase3}); a positive result would make the claim planner-invariant and a negative result would constrain it to lightweight heads.
\end{enumerate}

% ============================================================================
\section{Limitations}
\label{sec:concl-limitations}

\begin{itemize}
    \item \textbf{Open-loop evaluation only.} All results use PDMS/EPDMS, which are open-loop proxy metrics. Closed-loop simulation (e.g., in CARLA or nuPlan simulator) would provide stronger evidence of real-world generalization.

    \item \textbf{Limited geographic diversity.} NAVSIM/nuPlan covers only 4 cities (3 US + Singapore). Broader evaluation across more diverse geographies (Europe, developing countries, varied climate zones) is needed to fully validate the SSL generalization claim.

    \item \textbf{ImageNet pre-training.} Most SSL backbones were pre-trained on ImageNet, not driving data. While our nuScenes-pretrained I-JEPA shows competitive results, a comprehensive study of pre-training data domain is left for future work.

    \item \textbf{No weather/lighting variation.} Our cross-city protocol isolates geographic shift but does not independently test weather or lighting robustness, which is a complementary domain shift axis.

    \item \textbf{Camera-only constraint.} While motivated by edge deployment, removing LiDAR sacrifices 3D geometry information that may be critical in certain scenarios (e.g., occluded pedestrians, narrow alleys).
\end{itemize}

% ============================================================================
\section{Future Work}
\label{sec:concl-future}

\subsection{Video SSL Backbones (V-JEPA)}
The most immediate next step is to move from image-based SSL to video-based SSL. Driving is inherently temporal, so a predictive video objective should align more naturally with planning than single-frame pretraining. Drive-JEPA already suggests that V-JEPA-style pretraining can be especially strong in-distribution, which makes it a natural candidate for the next iteration of the cross-city robustness question.

\subsection{Closed-Loop Evaluation}
Closed-loop evaluation is the most important methodological extension. Although PDMS and EPDMS are stronger than raw displacement error, they remain proxy metrics. A closed-loop simulator such as CARLA or the nuPlan simulator would make it possible to test whether the same representation trends hold when the ego policy can alter future scene evolution.

\subsection{Driving-Specific Pre-Training at Scale}
Most backbones in the thesis still inherit generic large-scale pretraining. A larger study of driving-specific pretraining at scale would clarify how much of the cross-city robustness gain comes from the SSL objective itself and how much comes from domain alignment between pretraining and deployment.

\subsection{Scaling to More Cities and Datasets}
The current city-split benchmark covers only four cities and one data pipeline. Extending the same protocol to other datasets, and especially to cross-dataset transfer, would test whether the representation findings generalize beyond NAVSIM and nuScenes.

\subsection{RL Post-Training for Cross-City Transfer}
RQ6 remains open. If reward-guided post-training improves held-out-city behavior, then SSL + diffusion + RL becomes a plausible full recipe for robust planning. If it fails to transfer, then the next problem is reward design: current reward signals may be encoding city-specific structure rather than general driving competence.

\subsection{Multi-Modal JEPA Encoders}
Finally, the thesis motivates a multi-modal JEPA direction. A shared latent-space pretraining objective over camera and LiDAR could preserve the robustness benefits of SSL while recovering part of the geometry lost in camera-only deployment settings.
